Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents

### 2025-05-13 doc/cicm25-paper/paper.tex
ednots

Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both).

Concretely this is about converting the {\LaTeX} string
    (for conciseness completely unannotated)
    \ednote{MK: There is/was a discrepancy
  in the last line of the example; I tried to fix it, but failed\\
    FS: Is it fixed now? I don't see any discrepancy.}
\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item:
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
In the latter, the author has the word ``terminal'' with the concept it refers to with 
via the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.

\begin{newpart}{MK: more definitions}
  In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}:
words of the domain jargon used to refer to the concept or object.
Verbalizations can vary between between subdomains and communities of practice. In
our example, the verbalization ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation.

If the symbol name and the desired verbalization coincide (as they often do in English),
the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).\ednote{
    FS: Do we need the last paragraph?
        I've removed the ``sn utility'' annotation from the listing
        because I thought it breaks the flow a bit.
        But if we want to show the sn macro, I can add it back.
}
\end{newpart}

The annotation refers to a definition in the domain model module shown in
\cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in
line 3.

>>>>

Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to
annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both). Concretely this is about converting the {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item:
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
In the latter, the author has annotated the word ``terminal'' with the concept it refers
to with via the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.

In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
object.  Verbalizations can vary between between subdomains and communities of
practice. In our example, the verbalization ``terminal [state]'' was annotated with the
symbol with name \lstinline|goal state|. It comes from the module with name
\lstinline|search-problem| in the namespace \lstinline|smglom/search/mod| imported via the
\lstinline|\importmodule| macro in line 1. As the symbol name \lstinline|goal state| is
unique in the imports here it suffices as a relative URI in the \lstinline|\sr|
annotation. If the symbol name and the desired verbalization coincide (as they often do in
English), the short-hand macro \lstinline|\sn| can be used (e.g. as
\lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).

Both variants of the annotation refer to a definition in the domain model module shown in
\cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in
line 3.
